---
title: "Code collaboration in NEST"
author:
  - name: Ilse Augustyns
  - name: Nicholas Masel
  - name: Tad Lewandowski
  - name: Gabriel Becker
  - name: Clinical & Statistical Programming, IDAR, Johnson & Johnson, Innovative Medicine
description: "Announcement on code collaboration for the NEST framework between Roche and Clinical & Statistical Programming, Johnson & Johnson Innovative Medicine"
date: "2025-03-11"
# please do not use any non-default categories.
# You can find the default categories in the repository README.md
categories: [TLG, Community]
# feel free to change the image
image: "rtables_logo.svg"

---

<!--------------- typical setup ----------------->

```{r setup, include=FALSE}
long_slug <- "2025-03-11_code_collaborati..."
library(link)
link::auto()
```

<!--------------- post begins here ----------------->

In the ever-evolving field of clinical data analysis, innovation and collaboration are key. Today, we are pleased to share an initial, yet important milestone as part of a collaboration effort between Roche and Clinical & Statistical Programming, IDAR, J&J Innovative Medicine (further referred to as C&SP IDAR) within the NEST framework.

### Table creation in clinical reporting in pharma

In clinical trials, the presentation of data is as crucial as the data itself. Clear, accurate, and standardized tables ensure that the results are interpreted correctly and that regulatory bodies can easily review and approve submissions. 

For an overview of creation of tables within end-to-end clinical reporting in pharma, both [Pharmaverse TLGs Overview](https://pharmaverse.org/e2eclinical/tlg/) and the effort from [R Consortium](https://rconsortium.github.io/rtrs-wg/) are a useful read to compare examples from several packages.

### Code collaboration

Moving forward at C&SP IDAR, we are transitioning towards adopting R as our preferred analytical tool, and we were thrilled with the capabilities within the NEST framework ({rtables}, {tern} and {teal}) designed by Roche. 

As of today, we are building our new internal standard table shells using the open source packages from this NEST framework.

The {rtables} package is designed to simplify and standardize the process of creating tables for clinical trial reports. It offers a flexible and robust framework for generating high-quality, reproducible tables that adhere to industry standards. 

While the framework is robust in terms of type of analysis that can be performed, a key requirement for `C&SP IDAR` was not met. Our table output format requirement is the usage of `rtf` with `non-monospace` fonts.

As such, the initial scope of our collaboration was focusing on this **Output Format Customization** requirement. 

We are proud to share the first result of the collaboration effort! With the current [CRAN](https://cran.r-project.org/web/packages/rtables/index.html){target="_blank"} release of {rtables}, functions related to pagination or export can deal with `truetype` fonts. 
See the [NEWS updates](https://cran.r-project.org/web/packages/rtables/news/news.html){target="_blank"} (rtables 0.6.8) to find out more details on the new features.

While a formal `rtf` exporter function is not yet available, these updates were key to be able to create tables in our desired output format.
Further development will be done to release this functionality for broader usage.

By improving the efficiency and reliability of data presentation, researchers can focus more on the science and less on the formatting. With this update, we have to focus even less on output formatting issues.

### Get Involved

We believe in the power of community and open-source development. 

We are thankful to Roche for welcoming us as a new collaborator and giving us the opportunity to being part of this journey. Let's make clinical data analysis smarter, faster, and more reliable than ever before!

This partnership marks a significant milestone in our commitment to contribute to open source package development within a cross pharma network of companies. 

We are excited about the potential of NEST packages to transform the way clinical trial data is presented and are confident that, with combined expertise, our industry can achieve remarkable results.

For more information, visit the [NEST project page](https://insightsengineering.github.io/nest/index.html){target="_blank"}, and don't forget to follow our progress on [GitHub](https://github.com/insightsengineering){target="_blank"}.

As we embarked on this journey, we invite you to join us. Whether you’re an experienced user with feedback, a developer interested in contributing code, or a researcher looking to leverage {rtables}, {tern}, or {teal} for your projects, your input is invaluable.

We welcome both individual contributors and organizations. Please reach out to contribute via our [GitHub](https://insightsengineering.github.io/rtables/){target="_blank"}, via [pharmaverse/slack](https://pharmaverse.slack.com/)

<!--------------- appendices go here ----------------->

```{r, echo=FALSE, message=FALSE}
source("appendix.R")
insert_appendix(
  repo_spec = "pharmaverse/blog",
  name = long_slug,
  # file_name should be the name of your file
  file_name = list.files() %>% stringr::str_subset(".qmd") %>% first()
)
```
